\documentclass[11pt,oneside]{article}

% Centralized packages/macros
\input{other_tex/format_macros.tex}

% Optional math + theorem layers (safe if absent/empty)
\IfFileExists{other_tex/math.tex}{\input{other_tex/math.tex}}{}
\IfFileExists{other_tex/theorems.tex}{\input{other_tex/theorems.tex}}{}

% -------------
% Metadata
% -------------
\title{\PaperTitleMain}
\ifdefined\Authors
  \author{\Authors}
\else\ifdefined\AuthorName
  \author{\AuthorName}
\else
  \author{Author Name}
\fi\fi
\date{\today}

\begin{document}
\maketitle

% Dedication inline under the title (no page break)
\IfFileExists{other_tex/dedication.tex}{%
  \vspace{0.75\baselineskip}\input{other_tex/dedication.tex}%
}{}

% ----------------
% Abstract (single source of truth; add to ToC once)
% ----------------
\section*{Abstract}
\addcontentsline{toc}{section}{Abstract}
\IfFileExists{other_tex/abstract.tex}{\input{other_tex/abstract.tex}}{}

% =========================================================
% =====================  MAIN TEXT  =======================
% =========================================================

\section{Thermal Physics in the Planck-Cell Medium}
\subsection*{\textit{Why entropy per tick is the default language of change}}

% --- Hub/standalone-safe macro & notation guards ---
\providecommand{\Sent}{S_{\mathrm{ent}}}             % dimensionless entropy (nats)
% \providecommand{\chiRate}{\chi}                    % (removed to avoid symbol collision with \chi=S/t)
\newcommand{\Sphys}{S_{\mathrm{phys}}}               % entropy (J/K)
\newcommand{\dotSphys}{\dot{S}_{\mathrm{phys}}}      % entropy rate (J/K/s)

\section*{Scope}
Thermodynamics is often taught as the ``boiler room'' of physics—engines, refrigerators, heat transfer—while mechanics and fields get to be ``fundamental.'' In the Planck-Cell (S/t) framework, each tick \emph{permits} entropy exchange/production. Thermodynamics is therefore the \emph{default language of change}; mechanics and fields are corollaries of how entropy accumulates and flows \cite{callen1985thermodynamics}. Foundations of the tick-based framing appear in earlier parts of the program \cite{langstaff_zenodo_16908311}.

\emph{Scope guard.} Ticks supply \emph{time orientation} (before/after). Actual entropy growth follows from the local balance with nonnegative production in closed systems; cosmological low-entropy initial conditions are assumed, not derived here \cite{zeh2007arrow}.

% (Optional, put in your macros file to have a symbol for bits-entropy)
\newcommand{\Sbits}{S_{\mathrm{bits}}}

\subsection*{Notation \& units}
\begin{itemize}
  \item \textbf{Dimensioned entropy variables.}
  We use \(\Sphys\) (J/K) and \(\dotSphys\) (J/K/s) for all heat/temperature relations:
  \[
    \delta Q = T\,d\Sphys, \qquad \dotSphys = \frac{\dot Q}{T}.
  \]

  \item \textbf{Dimensionless counting.}
  We use \(\Sent\) (nats), with the rate identity
  \[
    \dot{\Sent} = \frac{\dotSphys}{k_B}.
  \]
  Conversion to bits uses
  \[
    S_{\text{bits}} = \frac{\Sent}{\ln 2}
    \quad\text{(or }\Sbits = \Sent/\ln 2\text{ if you enabled the macro).}
  \]

  \item \textbf{Not used here.}
  We \emph{do not} use the mechanics/gravitation potential-like field \(\chi = S/t\) from earlier chapters.
\end{itemize}


\paragraph{Units crosswalk (info lens).}
\[
  S_{\text{bits}} = \frac{\Sent}{\ln 2},
  \qquad
  \text{bits/s} = \frac{\dotSphys}{k_B \ln 2}.
\]

\paragraph{Planck-units checkpoint.}
\[
\begin{aligned}
\ell_P &= \sqrt{\frac{\hbar G}{c^{3}}} && [\mathrm{m}],\\
t_P    &= \sqrt{\frac{\hbar G}{c^{5}}} && [\mathrm{s}],\\
m_P    &= \sqrt{\frac{\hbar c}{G}}     && [\mathrm{kg}],\\
E_P    &= m_P c^{2}
        \;=\; \sqrt{\frac{\hbar c^{5}}{G}} && [\mathrm{J}],\\
T_P    &= \frac{E_P}{k_B}
        \;=\; \sqrt{\frac{\hbar c^{5}}{G}}\,\frac{1}{k_B} && [\mathrm{K}].
\end{aligned}
\]
\noindent\textit{Status:}
\(c\) and \(\hbar\) are fixed by kinematics/phase \cite{einstein1905elektrodynamik};
\(G\) by gravitation; and here \(k_B\) is anchored operationally via the heat–entropy
rate relation \(\dot S_{\mathrm{phys}}=\dot Q/T\) \cite{callen1985thermodynamics}.
With these identifications the temperature scale \(T_P\) is fixed (numerical values from
\cite{codata2018}).

% --------------------------
\section{Temperature as Update Rate}

\paragraph{Conventional view.}
Temperature measures average kinetic energy (in simple models) and serves as the integrating factor for heat \cite{boltzmann1872,callen1985thermodynamics}.

\paragraph{S/t view (rate-normalized entropy update).}
\[
\dotSphys \;=\; \frac{\dot Q}{T}
\qquad\Longleftrightarrow\qquad
T \;=\; \frac{\dot Q}{\dotSphys}.
\]
A hotter region is one with larger \(\dotSphys\).
Equilibrium means \emph{update rates equalize}—the Zeroth Law in rate language.

\paragraph{Tick calibration (link to nats and SI).}
Let a tick carry entropy \(s_{\rm tick}\) (J/K) and last \(\tau\). Then
\[
\dotSphys \;=\; \frac{s_{\rm tick}}{\tau},
\qquad
T \;=\; \frac{\dot Q\,\tau}{s_{\rm tick}}\,,
\qquad
\dot{\Sent}=\frac{\dotSphys}{k_B}.
\]

% --------------------------
\section{From Steam Engines to the Cosmos}

\paragraph{The steam engine.}
Nineteenth-century engines exposed the universality of entropy: some ticks are always consumed irreversibly. The classical laws were codified to describe those constraints \cite{clausius1879mechanical,callen1985thermodynamics}.

\paragraph{The cosmic engine.}
On the largest scales, the \emph{cosmic microwave background (CMB)} encodes the same logic. Early-universe ``ticking'' produced nearly scale-invariant fluctuations with a slight red tilt (\emph{slow-roll estimate} \(n_s\!\approx\!1-2/N\) with \(N\!\sim\!60\); observed \(n_s\!\approx\!0.965\)) \cite{peebles1993principles,planck2018params}.

\paragraph{Punchline (explicit equality under steady conditions).}
Over an interval where \(\dot Q\) and \(T\) are roughly constant,
\[
\Delta \Sphys \;\approx\; \frac{\dot Q}{T}\,\Delta t\,.
\]

% --------------------------
\section{Microstates, Multiplicity, and Statistical Temperature}

\paragraph{(1) Statistical origin.}
A macrostate corresponds to \(\Omega\) microstates:
\[
\Sent \;=\; \ln\Omega\quad\text{(nats)},\qquad
\Sphys \;=\; k_B\,\Sent\quad\text{(J/K)}.
\]
(For bits, divide by \(\ln 2\). Information theory mirrors this structure \cite{shannon1948mathematical}.)

\paragraph{(2) Temperature as sensitivity.}
\[
\frac{1}{T} \;=\; \left(\frac{\partial \Sphys}{\partial U}\right)_V .
\]

\paragraph{(3) Reservoir contact.}
Maximizing total multiplicity of system+\!bath gives \(\delta Q=T\,d\Sphys\): entropy gradients bias energy flow \cite{callen1985thermodynamics}.

\paragraph{(4) Local and global entropy balance.}
Let \(s\) be entropy density (J/K/m\(^3\)) and \(\mathbf j_S\) the entropy flux. Then
\[
\partial_t s \;+\; \nabla\!\cdot\!\mathbf j_S \;=\; \sigma\ \ (\sigma\ge 0),
\qquad
\mathbf j_S\;=\;\frac{\mathbf j_Q}{T}\;+\;s\,\mathbf v\;-\;\sum_i \frac{\mu_i}{T}\,\mathbf j_i,
\]
where \(\mathbf j_Q\) is heat flux, \(\mathbf v\) a bulk velocity, \(\mu_i\) chemical potentials and \(\mathbf j_i\) species fluxes (viscous terms contribute positively to \(\sigma\) and are omitted here for brevity) \cite{callen1985thermodynamics}. Integrated over a volume:
\[
\dot{S}_{\mathrm{phys}} \;=\; \sum_i \frac{\dot Q_i}{T_i} \;+\; \sigma.
\]

% --------------------------
\section{Heat and Internal Energy}

\paragraph{Heat.}
Heat \(Q\) is not a substance; it is transfer—path-dependent tick-generated updates between regions \cite{callen1985thermodynamics}.

\paragraph{Internal energy.}
\(U\) is the stored energy of microscopic updates (motions, configurations, fields)—a state function. Every degree of freedom is a tick source.

% --------------------------
\section{Heat Transfer Mechanisms}

\begin{itemize}
  \item \textbf{Conduction (diffusive updates).}
  Fourier’s law \(\mathbf j_Q=-\kappa\nabla T\) yields nonnegative entropy production:
  \[
    \dot{S}_{\mathrm{phys,cond}}=\int \mathbf j_Q\!\cdot\nabla(1/T)\,dV
    = \int \frac{\kappa\,|\nabla T|^2}{T^2}\,dV \;\ge\; 0\quad(\kappa>0).
  \]
  \item \textbf{Convection (bulk carriage).}
  Parcels move bodily; the entropy-flux contribution is \( \mathbf j_S^{\rm conv} = s\,\mathbf v \).
  \item \textbf{Radiation (ballistic updates).}
  Massless excitations carry updates one hop per tick, bounded by \(c\) \cite{einstein1905elektrodynamik,planck1901law}.
  For \emph{near-equilibrium blackbody-like} emission into vacuum,
  \[
    \dot{S}_{\mathrm{phys,rad}}=\frac{4}{3}\,\frac{P}{T}\,,
  \]
  with departures for non-Planckian spectra or reabsorption.
\end{itemize}

% --------------------------
\section{The Laws of Thermodynamics, Restated}

\subsection*{Zeroth Law}
Equal entropy-update rates \(\Rightarrow\) equal temperature. Thermal equilibrium is tick-rate synchronization.

\subsection*{First Law}
\textbf{Energy is conserved:}
\[
\Delta U \;=\; Q - W.
\]
\paragraph{Work from bias \(\times\) displacement (sign).}
We use \( \Delta U = Q - W\) with \(W\) the work \emph{by} the system:
\[
W \;=\; \int \mathbf F_{\rm by}\!\cdot d\mathbf x,\qquad
P \;=\; \mathbf F_{\rm by}\!\cdot \mathbf v.
\]
(Equivalently, with external work on the system, \(W_{\rm on}=-W\) and \(\Delta U=Q+W_{\rm on}\).)

\subsection*{Clausius inequality}
For any process between fixed end states,
\[
\Delta \Sphys \;\ge\; \int_{t_0}^{t_1} \frac{\delta Q_{\rm rev}}{T},
\]
with equality for reversible paths; for any cycle,
\[
\oint \frac{\delta Q}{T} \;\le\; 0,
\]
with equality if reversible \cite{clausius1879mechanical,callen1985thermodynamics}.

\subsection*{Second Law}
For a \emph{closed system}, the global entropy is non-decreasing; locally, production satisfies \(\sigma\ge 0\).
Entropy can decrease in a \emph{subsystem} only by exporting it via flux \(\mathbf j_S\) \cite{boltzmann1872,shannon1948mathematical,zeh2007arrow}.

\subsection*{Third Law}
If local ticking halts, entropy approaches a constant minimum. For a perfect crystal, this limit is conventionally zero. In the Planck-Cell framing: no ticks, no entropy production.

% --------------------------
\section{Worked Examples}

\subsection*{1) Steam Engine (macro)}
A heat engine takes in \(Q_H\) at \(T_H\) and exhausts \(Q_C\) at \(T_C\).
Carnot efficiency:
\[
\eta \;=\; 1 - \frac{T_C}{T_H}.
\]
Each hot tick carries \(\Delta \Sphys = Q_H/T_H\); running requires dumping \(\Delta S_{\mathrm{phys,C}} = Q_C/T_C\).
The piston is powered by the update-rate difference: an \emph{entropy-tick converter} \cite{callen1985thermodynamics}.

\subsection*{2) A Cup of Water (everyday \(\dotSphys\)-meter)}
Heat \(1~\text{kg}\) of water from \(300\ \text{K}\) to \(310\ \text{K}\) in \(100\ \text{s}\), with \(c_p = 4180\ \text{J/(kg\,K)}\) \cite{crc_handbook}.
\[
\Delta \Sphys \approx m c_p \ln\!\frac{310}{300} \approx 1.37\times 10^{2}\ \text{J/K},\qquad
\dotSphys \approx 1.37 \ \text{J/(K\,s)}.
\]

\subsection*{3) CMB (cosmic)}
Early-universe ticking produced nearly scale-invariant fluctuations with red tilt
\(n_s \approx 1 - 2/N\) for \(N \sim 60\) (\emph{slow-roll} heuristic; observed \(n_s\!\approx\!0.965\)) \cite{peebles1993principles,planck2018params}.

\subsection*{4) The Sun as a Cosmic \(\dotSphys\)-meter}
For near–blackbody radiation at temperature \(T\),
\[
\dot{S}_{\mathrm{phys,rad}} \;=\; \frac{4}{3}\,\frac{P}{T}.
\]
With \(L_\odot \simeq 3.85\times 10^{26}\ \mathrm{W}\) and \(T_\odot \simeq 5.77\times 10^3\ \mathrm{K}\) \cite{allen2000},
\[
\dot{S}_{\mathrm{phys},\odot} \;\approx\; \frac{4}{3}\,\frac{L_\odot}{T_\odot}
\;\approx\; 8.9\times 10^{22}\ \mathrm{J\,K^{-1}\,s^{-1}}.
\]
Absorption by space at \(T_{\rm space}\approx 2.725\ \mathrm{K}\) \cite{planck2018params} gives the net production
\[
\dot{S}_{\mathrm{phys,univ}} \;\approx\; \frac{4}{3}\,L_\odot\!\left(\frac{1}{T_{\rm space}} - \frac{1}{T_\odot}\right)
\;\approx\; 1.9\times 10^{26}\ \mathrm{J\,K^{-1}\,s^{-1}}.
\]
\emph{Bits/s (info lens):}
\[
\frac{\dot{S}_{\mathrm{phys},\odot}}{k_B\ln 2}\ \approx\ 9.3\times 10^{45}\ \text{bits/s},\qquad
\frac{\dot{S}_{\mathrm{phys,univ}}}{k_B\ln 2}\ \approx\ 2.0\times 10^{49}\ \text{bits/s}.
\]
\footnotesize\noindent\emph{Constants used (standard values):} CODATA fundamental constants \cite{codata2018}; solar luminosity and photospheric temperature \cite{allen2000}; water heat capacity \cite{crc_handbook}.
\normalsize

% --------------------------
\section{Interpretation and Foreshadow}
Thermodynamics is not about engines; it is about ticks.
Every law of thermal physics flows from the S/t axiom:
\[
\text{Ticks bias systems toward increasing entropy (\(\sigma\ge 0\)).}
\]
Temperature, heat, and the four laws are systematic bookkeeping for those ticks.
Later we connect the same rate picture to the CMB spectrum, tilt, and polarization—seamlessly from kettles to cosmos \cite{peebles1993principles,planck2018params}. For the broader Planck-Cell program and axioms, see prior articles by the author \cite{langstaff_zenodo_16908311}.

% --------------------------
\section*{Summary box}
\begin{center}
\begingroup
\setlength{\fboxsep}{9pt}% inner padding
\setlength{\fboxrule}{0.6pt}% border thickness
\fbox{%
  \begin{minipage}{0.95\linewidth}
    \setlength{\parindent}{0pt}
    \raggedright\linespread{1.08}\selectfont
    \textbf{Key relations and statements}\par\medskip
    \begin{itemize}
      \setlength{\itemsep}{0.9ex}
      \setlength{\topsep}{0.6ex}
      \setlength{\parsep}{0.4ex}
      \setlength{\partopsep}{0pt}
      \item \textbf{Temperature as update rate:} \(\dotSphys=\dot Q/T \Rightarrow T=\dot Q/\dotSphys\); \(\dot{\Sent}=\dotSphys/k_B\).
      \item \textbf{State vs.\ path:} Heat is path-dependent; internal energy \(U\) is a state function.
      \item \textbf{Microstate entropy:} \(\Sent=\ln\Omega\) (nats), \(\Sphys=k_B\Sent\); information rate: \(\text{bits/s}=\dot{S}_{\mathrm{phys}}/(k_B\ln 2)\).
      \item \textbf{Local balance:} \( \partial_t s+\nabla\!\cdot\!\mathbf j_S=\sigma\), with \( \mathbf j_S=\mathbf j_Q/T + s\,\mathbf v - \sum_i (\mu_i/T)\,\mathbf j_i\), and \(\sigma\ge 0\).
      \item \textbf{Global balance:} \( \dot{S}_{\mathrm{phys}}=\sum_i \dot Q_i/T_i+\sigma\); conduction (\(\kappa>0\)) \(\Rightarrow\ \dot{S}_{\mathrm{phys,cond}}\ge 0\).
      \item \textbf{Radiation (near-BB):} \(\dot{S}_{\mathrm{phys,rad}}=\tfrac{4}{3}\,P/T\) (vacuum; reabsorption/non-Planckian spectra modify this).
      \item \textbf{First Law:} \(\Delta U = Q - W\), with \(W=\int \mathbf F_{\rm by}\!\cdot d\mathbf x\) and \(P=\mathbf F_{\rm by}\!\cdot \mathbf v\).
      \item \textbf{Clausius statements:} \(\Delta \Sphys \ge \int \delta Q_{\rm rev}/T\); \(\oint \delta Q/T \le 0\) (equality if reversible).
      \item \textbf{Meters:} Heat engines as entropy–tick converters; Sun/CMB as near-blackbody \(\dotSphys\)-meters.
    \end{itemize}
  \end{minipage}%
}
\endgroup
\end{center}

% =========================================================
% ===============  BACK MATTER / STATEMENTS  ==============
% =========================================================

\clearpage
\section*{Acknowledgments}
\addcontentsline{toc}{section}{Acknowledgments}
\IfFileExists{other_tex/acknowledgments.tex}{\input{other_tex/acknowledgments.tex}}{}

\section*{Competing interests}
\IfFileExists{other_tex/competing_interests.tex}{\input{other_tex/competing_interests.tex}}{}

\section*{Author contributions}
\IfFileExists{other_tex/authors_contributions.tex}{\input{other_tex/authors_contributions.tex}}{}

\section*{Funding}
\IfFileExists{other_tex/funding.tex}{\input{other_tex/funding.tex}}{}

\section*{Data and materials availability}
\IfFileExists{data_accessibility.tex}{\input{data_accessibility.tex}}{No new data were generated.}

\section*{Disclosure}
\IfFileExists{other_tex/disclosure.tex}{\input{other_tex/disclosure.tex}}{}

% ----------------
% Other Works (inserted above bibliography)
% ----------------
\clearpage
\section*{Other Works by the Author}
\addcontentsline{toc}{section}{Other Works by the Author}
The present article on thermal physics in the Planck-Cell medium builds on a sequence
of related works that establish the framework step by step:
\begin{itemize}
  \item \textbf{Temporal Relativity.} C.\,M. Langstaff (2025). Zenodo. DOI: 10.5281/zenodo.17119049.
  \item \textbf{Planck-Cell Kinematics.} C.\,M. Langstaff (2025). Zenodo. DOI: 10.5281/zenodo.17168478.
  \item \textbf{Planck-Cell Mass.} C.\,M. Langstaff (2025). Zenodo. DOI: 10.5281/zenodo.17209646.
  \item \textbf{Planck-Cell Gravity.} C.\,M. Langstaff (2025). Zenodo. DOI: 10.5281/zenodo.17210232.
\end{itemize}
Together, these works define the axioms, kinematic rules, and stiffness
interpretation of mass in the Planck-Cell medium, providing the foundation
on which the present thermal layer is constructed.

% ----------------
% Glossary (prints only if glossaries is loaded)
% ----------------
\IfGlossariesEnabled{
  \clearpage
  \printglossary[title=Glossary]
}

% ----------------
% Bibliography (BibTeX)
% ----------------
\clearpage
\bibliographystyle{plain}
% (No \nocite{*}; only cited items will appear.)
\bibliography{bibliography}

\end{document}
