% --- Hub/standalone-safe macro & notation guards ---
\providecommand{\Sent}{S_{\mathrm{ent}}}             % dimensionless entropy (nats)
% \providecommand{\chiRate}{\chi}                    % (removed to avoid symbol collision with \chi=S/t)
\newcommand{\Sphys}{S_{\mathrm{phys}}}               % entropy (J/K)
\newcommand{\dotSphys}{\dot{S}_{\mathrm{phys}}}      % entropy rate (J/K/s)

\section*{Temperature as Update Rate}

\paragraph{Conventional view.}
Temperature measures average kinetic energy (in simple models) and serves as the integrating factor for heat \cite{boltzmann1872,callen1985thermodynamics}.

\paragraph{S/t view (rate-normalized entropy update).}
\[
\dotSphys \;=\; \frac{\dot Q}{T}
\qquad\Longleftrightarrow\qquad
T \;=\; \frac{\dot Q}{\dotSphys}.
\]
A hotter region is one with larger \(\dotSphys\).
Equilibrium means \emph{update rates equalize}—the Zeroth Law in rate language.

\paragraph{Tick calibration (link to nats and SI).}
Let a tick carry entropy \(s_{\rm tick}\) (J/K) and last \(\tau\). Then
\[
\dotSphys \;=\; \frac{s_{\rm tick}}{\tau},
\qquad
T \;=\; \frac{\dot Q\,\tau}{s_{\rm tick}}\,,
\qquad
\dot{\Sent}=\frac{\dotSphys}{k_B}.
\]

% --------------------------
\section*{From Steam Engines to the Cosmos}

\paragraph{The steam engine.}
Nineteenth-century engines exposed the universality of entropy: some ticks are always consumed irreversibly. The classical laws were codified to describe those constraints \cite{clausius1879mechanical,callen1985thermodynamics}.

\paragraph{The cosmic engine.}
On the largest scales, the \emph{cosmic microwave background (CMB)} encodes the same logic. Early-universe ``ticking'' produced nearly scale-invariant fluctuations with a slight red tilt (\emph{slow-roll estimate} \(n_s\!\approx\!1-2/N\) with \(N\!\sim\!60\); observed \(n_s\!\approx\!0.965\)) \cite{peebles1993principles,planck2018params}.

\paragraph{Punchline (explicit equality under steady conditions).}
Over an interval where \(\dot Q\) and \(T\) are roughly constant,
\[
\Delta \Sphys \;\approx\; \frac{\dot Q}{T}\,\Delta t\,.
\]

% --------------------------
\section*{Microstates, Multiplicity, and Statistical Temperature}

\paragraph{(1) Statistical origin.}
A macrostate corresponds to \(\Omega\) microstates:
\[
\Sent \;=\; \ln\Omega\quad\text{(nats)},\qquad
\Sphys \;=\; k_B\,\Sent\quad\text{(J/K)}.
\]
(For bits, divide by \(\ln 2\). Information theory mirrors this structure \cite{shannon1948mathematical}.)

\paragraph{(2) Temperature as sensitivity.}
\[
\frac{1}{T} \;=\; \left(\frac{\partial \Sphys}{\partial U}\right)_V .
\]

\paragraph{(3) Reservoir contact.}
Maximizing total multiplicity of system+\!bath gives \(\delta Q=T\,d\Sphys\): entropy gradients bias energy flow \cite{callen1985thermodynamics}.

\paragraph{(4) Local and global entropy balance.}
Let \(s\) be entropy density (J/K/m\(^3\)) and \(\mathbf j_S\) the entropy flux. Then
\[
\partial_t s \;+\; \nabla\!\cdot\!\mathbf j_S \;=\; \sigma\ \ (\sigma\ge 0),
\qquad
\mathbf j_S\;=\;\frac{\mathbf j_Q}{T}\;+\;s\,\mathbf v\;-\;\sum_i \frac{\mu_i}{T}\,\mathbf j_i,
\]
where \(\mathbf j_Q\) is heat flux, \(\mathbf v\) a bulk velocity, \(\mu_i\) chemical potentials and \(\mathbf j_i\) species fluxes (viscous terms contribute positively to \(\sigma\) and are omitted here for brevity) \cite{callen1985thermodynamics}. Integrated over a volume:
\[
\dot{S}_{\mathrm{phys}} \;=\; \sum_i \frac{\dot Q_i}{T_i} \;+\; \sigma.
\]

% --------------------------
\section*{Heat and Internal Energy}

\paragraph{Heat.}
Heat \(Q\) is not a substance; it is transfer—path-dependent tick-generated updates between regions \cite{callen1985thermodynamics}.

\paragraph{Internal energy.}
\(U\) is the stored energy of microscopic updates (motions, configurations, fields)—a state function. Every degree of freedom is a tick source.

% --------------------------
\section*{Heat Transfer Mechanisms}

\begin{itemize}
  \item \textbf{Conduction (diffusive updates).}
  Fourier’s law \(\mathbf{j}_Q=-\kappa\nabla T\) gives nonnegative entropy production:
  \begin{equation}
  \begin{aligned}
  \dot{S}_{\mathrm{phys,cond}}
    &= \int \mathbf{j}_Q \!\cdot \nabla(1/T)\,\mathrm{d}V \\
    &= \int \frac{\kappa\,\lVert\nabla T\rVert^2}{T^2}\,\mathrm{d}V
       \;\ge\; 0 \qquad (\kappa>0).
  \end{aligned}
  \end{equation}

  \item \textbf{Convection (bulk carriage).}
  Parcels move bodily; the entropy–flux contribution is
  \( \mathbf{j}_S^{\mathrm{conv}} = s\,\mathbf{v} \).

  \item \textbf{Radiation (ballistic updates).}
  Massless excitations carry updates one hop per tick, bounded by \(c\) \cite{einstein1905elektrodynamik,planck1901law}.
  For \emph{near-equilibrium, blackbody-like} emission into vacuum,
  \[
    \dot{S}_{\mathrm{phys,rad}}=\frac{4}{3}\,\frac{P}{T}\,,
  \]
  with departures for non-Planckian spectra or reabsorption.
\end{itemize}


% --------------------------
\section*{The Laws of Thermodynamics, Restated}

\subsection*{Zeroth Law}
Equal entropy-update rates \(\Rightarrow\) equal temperature. Thermal equilibrium is tick-rate synchronization.

\subsection*{First Law}
\textbf{Energy is conserved:}
\[
\Delta U \;=\; Q - W.
\]
\paragraph{Work from bias \(\times\) displacement (sign).}
We use \( \Delta U = Q - W\) with \(W\) the work \emph{by} the system:
\[
W \;=\; \int \mathbf F_{\rm by}\!\cdot d\mathbf x,\qquad
P \;=\; \mathbf F_{\rm by}\!\cdot \mathbf v.
\]
(Equivalently, with external work on the system, \(W_{\rm on}=-W\) and \(\Delta U=Q+W_{\rm on}\).)

\subsection*{Clausius inequality}
For any process between fixed end states,
\[
\Delta \Sphys \;\ge\; \int_{t_0}^{t_1} \frac{\delta Q_{\rm rev}}{T},
\]
with equality for reversible paths; for any cycle,
\[
\oint \frac{\delta Q}{T} \;\le\; 0,
\]
with equality if reversible \cite{clausius1879mechanical,callen1985thermodynamics}.

\subsection*{Second Law}
For a \emph{closed system}, the global entropy is non-decreasing; locally, production satisfies \(\sigma\ge 0\).
Entropy can decrease in a \emph{subsystem} only by exporting it via flux \(\mathbf j_S\) \cite{boltzmann1872,shannon1948mathematical,zeh2007arrow}.

\subsection*{Third Law}
If local ticking halts, entropy approaches a constant minimum. For a perfect crystal, this limit is conventionally zero. In the Planck-Cell framing: no ticks, no entropy production.

% --------------------------
\section*{Worked Examples}

\subsection*{1) Steam Engine (macro)}
A heat engine takes in \(Q_H\) at \(T_H\) and exhausts \(Q_C\) at \(T_C\).
Carnot efficiency:
\[
\eta \;=\; 1 - \frac{T_C}{T_H}.
\]
Each hot tick carries \(\Delta \Sphys = Q_H/T_H\); running requires dumping \(\Delta S_{\mathrm{phys,C}} = Q_C/T_C\).
The piston is powered by the update-rate difference: an \emph{entropy-tick converter} \cite{callen1985thermodynamics}.

\subsection*{2) A Cup of Water (everyday \(\dotSphys\)-meter)}
Heat \(1~\text{kg}\) of water from \(300\ \text{K}\) to \(310\ \text{K}\) in \(100\ \text{s}\), with \(c_p = 4180\ \text{J/(kg\,K)}\) \cite{crc_handbook}.
\[
\Delta \Sphys \approx m c_p \ln\!\frac{310}{300} \approx 1.37\times 10^{2}\ \text{J/K},\qquad
\dotSphys \approx 1.37 \ \text{J/(K\,s)}.
\]

\subsection*{3) CMB (cosmic)}
Early-universe ticking produced nearly scale-invariant fluctuations with red tilt
\(n_s \approx 1 - 2/N\) for \(N \sim 60\) (\emph{slow-roll} heuristic; observed \(n_s\!\approx\!0.965\)) \cite{peebles1993principles,planck2018params}.

\subsection*{4) The Sun as a Cosmic \(\dotSphys\)-meter}
For near–blackbody radiation at temperature \(T\),
\[
\dot{S}_{\mathrm{phys,rad}} \;=\; \frac{4}{3}\,\frac{P}{T}.
\]
With \(L_\odot \simeq 3.85\times 10^{26}\ \mathrm{W}\) and \(T_\odot \simeq 5.77\times 10^3\ \mathrm{K}\) \cite{allen2000},
\[
\dot{S}_{\mathrm{phys},\odot} \;\approx\; \frac{4}{3}\,\frac{L_\odot}{T_\odot}
\;\approx\; 8.9\times 10^{22}\ \mathrm{J\,K^{-1}\,s^{-1}}.
\]
Absorption by space at \(T_{\rm space}\approx 2.725\ \mathrm{K}\) \cite{planck2018params} gives the net production
\[
\dot{S}_{\mathrm{phys,univ}} \;\approx\; \frac{4}{3}\,L_\odot\!\left(\frac{1}{T_{\rm space}} - \frac{1}{T_\odot}\right)
\;\approx\; 1.9\times 10^{26}\ \mathrm{J\,K^{-1}\,s^{-1}}.
\]
\emph{Bits/s (info lens):}
\[
\frac{\dot{S}_{\mathrm{phys},\odot}}{k_B\ln 2}\ \approx\ 9.3\times 10^{45}\ \text{bits/s},\qquad
\frac{\dot{S}_{\mathrm{phys,univ}}}{k_B\ln 2}\ \approx\ 2.0\times 10^{49}\ \text{bits/s}.
\]
\footnotesize\noindent\emph{Constants used (standard values):} CODATA fundamental constants \cite{codata2018}; solar luminosity and photospheric temperature \cite{allen2000}; water heat capacity \cite{crc_handbook}.
\normalsize

% --------------------------
\section*{Interpretation and Foreshadow}
Thermodynamics is not about engines; it is about ticks.
Every law of thermal physics flows from the S/t axiom:
\[
\text{Ticks bias systems toward increasing entropy (\(\sigma\ge 0\)).}
\]
Temperature, heat, and the four laws are systematic bookkeeping for those ticks.
Later we connect the same rate picture to the CMB spectrum, tilt, and polarization—seamlessly from kettles to cosmos \cite{peebles1993principles,planck2018params}. For the broader Planck-Cell program and axioms, see prior articles by the author \cite{langstaff_zenodo_16908311}.

% --------------------------
\section*{Summary box}
\begin{center}
\begingroup
\setlength{\fboxsep}{9pt}% inner padding
\setlength{\fboxrule}{0.6pt}% border thickness
\fbox{%
  \begin{minipage}{0.95\linewidth}
    \setlength{\parindent}{0pt}
    \raggedright\linespread{1.08}\selectfont
    \textbf{Key relations and statements}\par\medskip
    \begin{itemize}
      \setlength{\itemsep}{0.9ex}
      \setlength{\topsep}{0.6ex}
      \setlength{\parsep}{0.4ex}
      \setlength{\partopsep}{0pt}
      \item \textbf{Temperature as update rate:} \(\dotSphys=\dot Q/T \Rightarrow T=\dot Q/\dotSphys\); \(\dot{\Sent}=\dotSphys/k_B\).
      \item \textbf{State vs.\ path:} Heat is path-dependent; internal energy \(U\) is a state function.
      \item \textbf{Microstate entropy:} \(\Sent=\ln\Omega\) (nats), \(\Sphys=k_B\Sent\); information rate: \(\text{bits/s}=\dot{S}_{\mathrm{phys}}/(k_B\ln 2)\).
      \item \textbf{Local balance:} \( \partial_t s+\nabla\!\cdot\!\mathbf j_S=\sigma\), with \( \mathbf j_S=\mathbf j_Q/T + s\,\mathbf v - \sum_i (\mu_i/T)\,\mathbf j_i\), and \(\sigma\ge 0\).
      \item \textbf{Global balance:} \( \dot{S}_{\mathrm{phys}}=\sum_i \dot Q_i/T_i+\sigma\); conduction (\(\kappa>0\)) \(\Rightarrow\ \dot{S}_{\mathrm{phys,cond}}\ge 0\).
      \item \textbf{Radiation (near-BB):} \(\dot{S}_{\mathrm{phys,rad}}=\tfrac{4}{3}\,P/T\) (vacuum; reabsorption/non-Planckian spectra modify this).
      \item \textbf{First Law:} \(\Delta U = Q - W\), with \(W=\int \mathbf F_{\rm by}\!\cdot d\mathbf x\) and \(P=\mathbf F_{\rm by}\!\cdot \mathbf v\).
      \item \textbf{Clausius statements:} \(\Delta \Sphys \ge \int \delta Q_{\rm rev}/T\); \(\oint \delta Q/T \le 0\) (equality if reversible).
      \item \textbf{Meters:} Heat engines as entropy–tick converters; Sun/CMB as near-blackbody \(\dotSphys\)-meters.
    \end{itemize}
  \end{minipage}%
}
\endgroup
\end{center}

%%% PATCH BEGIN: MS-PL-001
\section*{Clarifications}
\paragraph{Temporal stiffness formalization}
Let a local mode satisfy \(\partial_\tau^2\phi=-\omega_0^2\phi\). The dispersion \(\omega^2=c^2\kappa^2+\omega_0^2\) implies \(E^2=(pc)^2+(mc^2)^2\) with \(m=\hbar\omega_0/c^2\).
\paragraph{Species-local vs universal}
\(\omega_0\) is species-local. A practical bound: \(|\dot\omega_0/\omega_0|\) constrained by precision clock comparisons.
\paragraph{Anchors}
Compton, plasma, and Debye frequencies provide direct \(\omega_0\) readouts (list numeric examples in text).
%%% PATCH END: MS-PL-001

\paragraph{Units.}
\begin{description}
\item[$T$] K \quad \item[$\dot Q$] J s$^{-1}$
\item[$S_{\mathrm{phys}}$] J K$^{-1}$ \quad \item[$\dot S_{\mathrm{phys}}$] J K$^{-1}$ s$^{-1}$
\item[$s$] J K$^{-1}$ m$^{-3}$ \quad \item[$j_S$] J K$^{-1}$ m$^{-2}$ s$^{-1}$
\item[$\sigma$] J K$^{-1}$ m$^{-3}$ s$^{-1}$
\end{description}

\begin{table}[h]
\centering
\begin{tabular}{lll}
\toprule
Mechanism & Statement & Entropy rate \\
\midrule
Conduction & $j_Q=-\kappa \, \nabla T$ & $\dot S_{\mathrm{cond}}=\int \kappa \frac{(\nabla T)^2}{T^2}\,dV \ge 0$ \\
Convection & advect $s$ with velocity $v$ & $j_S^{\mathrm{conv}} = s\,v$ \\
Radiation (near-bb) & Poynting flux & $\dot S_{\mathrm{rad}} = \tfrac{4}{3}\,P/T$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Conclusion.}
In the tick picture, thermodynamic laws are rate bookkeeping: $\dot S_{\mathrm{phys}}=\dot Q/T$ and $\partial_t s+\nabla\!\cdot j_S=\sigma$ define temperature, flow, and the arrow ($\sigma\ge 0$) without extra postulates. This frames later papers---electromagnetism as a capacity fraction and expansion as global entropy creation---in the same language.