\section*{Notation}

\noindent\textbf{Conventions.}  
Time and entropy advance in discrete, \emph{dimensionless ticks}.  
A single primitive cell unit is created per tick, giving
\[
  \Delta S = \Delta \tau.
\]
Physical mapping is applied only when necessary:
\[
  S_{\mathrm{phys}} = k_B\,S, \qquad
  \tau_s = t_P\,\tau.
\]
Fundamental constants: speed of light \(c\), reduced Planck constant \(\hbar\), and Boltzmann constant \(k_B\).

\smallskip
\noindent\textit{Native--SI correspondence.}  
One hop per tick implies \(\mathrm{d}a/\mathrm{d}\tau = 1\).  
Choosing a length per hop \(\ell_0\) and a time per tick \(t_0\) yields
\[
  c = \frac{\ell_0}{t_0}.
\]

\bigskip
\noindent\textbf{Core quantities.}
\begin{description}[leftmargin=2.4em,labelsep=0.8em]
  \item[\(k\)] Tick index (integer). The scale factor \(a(k)\) is defined with respect to ticks; baseline \(a(k)\propto k^{1/3}\).
  \item[\(\Delta S,\,\Delta\tau\)] Primitive-cell and proper-time tick \emph{counts} (dimensionless) satisfying \(\Delta S=\Delta\tau\).
  \item[\(S_{\mathrm{phys}}\)] Physical entropy (J K\(^{-1}\)), with \(S_{\mathrm{phys}} = k_B S\).
  \item[\(\dot S_{\mathrm{phys}}\)] Entropy rate (J K\(^{-1}\) s\(^{-1}\)).
  \item[\(T\)] Temperature (K), with \(\dot S_{\mathrm{phys}} = \dot Q / T\) near equilibrium.
  \item[\(\dot Q\)] Heat flow rate (W = J s\(^{-1}\)).
  \item[\(s\)] Entropy density (J K\(^{-1}\) m\(^{-3}\)).
  \item[\(j_S\)] Entropy flux density (J K\(^{-1}\) m\(^{-2}\) s\(^{-1}\)).
  \item[\(\sigma\)] Local entropy production rate (J K\(^{-1}\) m\(^{-3}\) s\(^{-1}\)), satisfying
        \(\partial_t s + \nabla\!\cdot j_S = \sigma.\)
  \item[\(\chi\)] Entropy–time update rate, \(\chi := S/t\) (s\(^{-1}\)).
\end{description}

\noindent\textbf{Graph and geometric parameters.}
\begin{description}[leftmargin=2.4em,labelsep=0.8em]
  \item[\(d_G(x,y)\)] Graph (hop) distance between vertices \(x\) and \(y\).
  \item[\(\ell_N\)] Typical interior hop length for a sample of size \(N\).
  \item[\(\varepsilon_N,\,\delta_N\)] Mesh non-uniformity and directional-bias parameters (\(\to 0\) as \(N\to\infty\)).
  \item[\(\eta_N\)] Combined small parameter: \(\eta_N := C_1\varepsilon_N + C_2\delta_N.\)
  \item[\(\mathrm{distortion}(N)\)] Bilipschitz distortion of the graph metric relative to the ambient metric:
  \[
    \mathrm{distortion}(N)
      = \sup_{x,y}
        \max\!\left\{
          \frac{d_G(x,y)}{\|x-y\|},\,
          \frac{\|x-y\|}{d_G(x,y)}
        \right\}-1.
  \]
\end{description}

\noindent\textbf{Thermodynamics-specific quantities.}
\begin{description}[leftmargin=2.4em,labelsep=0.8em]
  \item[\(T,\,Q\)] Temperature and heat; thermodynamic entropy \(S_{\mathrm{th}} = k_B S\).
  \item[\(\dot S\)] In SI units, \(\dot S_{\mathrm{th}} = \dot Q / T\) corresponds to \(\dot S = \dot Q / (k_B T)\) in tick form.
  \item[\(s,\, j_S,\, \sigma\)] Entropy density, flux, and production rate defined through  
        \(\partial_t s + \nabla\!\cdot j_S = \sigma.\)
  \item[\(\chi\)] Entropy–time update rate, \(\chi := S/t\) (s\(^{-1}\)); sets the temporal scale for thermodynamic drift.
\end{description}

\bigskip
\noindent\textbf{Quick symbol summary.}
\begin{description}[leftmargin=2.4em,labelsep=0.8em]
  \item[\(\ell_P\)] Planck-cell edge or hop length.
  \item[\(t_P\)] Planck tick (one global update interval).
  \item[\(c=\ell_P/t_P\)] Causal speed cap (one hop per tick).
  \item[\(k\)] Discrete tick index (global ledger step).
  \item[\(R(k)\)] Front radius after \(k\) ticks.
  \item[\(N(k)\)] Number of active/born cells after \(k\) ticks.
  \item[\(L\)] Continuum baseline distance.
  \item[\(d_G\)] Graph (hop) distance.
  \item[\(T(L)\)] Transit time over baseline \(L\).
  \item[\(v_g\)] Group or front speed.
  \item[\(E,\,\hbar\)] Energy and reduced Planck constant.
  \item[\(\Delta\phi\)] Accumulated phase along the path.
  \item[\(\lambda\)] Wavelength of the probe signal.
\end{description}
